WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-07-08 ms.tex
Update parallelization and run time sections

\subsection{Integral over distance}

The innermost integral over distance\footnote{The distance integral is now performed differently from what we initially described in \cite{leo-singer-thesis}; the method described in the present work is about an order of magnitude faster.} $r$ may be written as
%
\begin{align}\label{eq:distance-integral}
    \mathscr{F} &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}\right]
        I_0\left[\frac{b}{r}\right] r^m dr \nonumber \\
    &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}+\frac{b}{r}\right]
        \overline{I}_0\left[\frac{b}{r}\right] r^m dr
\end{align}
%
or, completing the square,
%
\begin{align}
    \mathscr{F} &= \exp\left[\frac{p^2}{{r_0}^2}\right]
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\left(\frac{p}{r} - \frac{p}{r_0}\right)^2\right]
        \overline{I}_0\left[\frac{2p^2}{r r_0}\right] r^m dr \\
    &= \exp\left[\frac{p^2}{{r_0}^2}\right] \mathscr{G},
\end{align}
%
where
%
\begin{align}
    p^2 &= \frac{1}{2} r^2 \sum_i {\rho_i}^2 \label{eq:likelihood-p-factor} \\
    b &= r \left| \sum_i \hat\rho_i \rho_i e^{i \tilde{\gamma}_i}
        a_i^*(\tilde{\tau}_i) \right| \label{eq:likelihood-b-factor} \\
    r_0 &= 2 p^2 / b \\
    \overline{I}_0(x) &= \exp(-|x|) I_0(x).
\end{align}
%
The coefficients $p^2$ and $b$ are nonnegative and independent of distance. $p$ has a maximum value of
%
\begin{equation}
    p_\mathrm{max} = \sqrt{\frac{1}{2} \sum_i \left(\frac{r_{1,i}}{r}\right)^2}.
\end{equation}
%
The symbol $\overline{I}_0$ denotes an exponentially scaled Bessel function. In the limit of large argument, $I_0(|x|) \sim \exp(|x|) / \sqrt{2 \pi |x|}$ \citep{Olver:2010:NHMF,NIST:DLMF}\footnote{\url{http://dlmf.nist.gov/10.40.E1}}. The scaled Bessel is useful for evaluation on a computer because it has a relatively small range $(0, 1]$ and varies slowly in proportion to $x^{1/2}$.

@@

Within each iteration, all of the marginalization integrals may be evaluated in parallel. In our C language implementation of \ac{BAYESTAR}, the inner loop over pixels is parallelized with OpenMP\footnote{http://openmp.org/}.

@@

We measured sky localization areas for each event as follows. First, we ranked the \ac{HEALPix} pixels by descending posterior probability. Then, we computed the cumulative sum of the pixels in that order. Finally, we searched for the index of the pixel whose cumulative sum was equal to a given value: for example, 0.9 if we are interested in the 90\% credible area. That pixel index times the area per pixel is the area of the smallest region of the specified credible level.

We can construct a second measure, called the searched area, as the smallest such constructed area that contains the true location of the source. A telescope with a \ac{FOV} that is small compared to the characteristic scale of the posterior would intercept the true location of the source after covering the searched area. This measure is mainly useful because it measures the accuracy of the sky localization independently of the precision: that is to say, it treats the sky map as merely a ranking statistic.

@@

\subsection{Self-consistency}

\textbf{need to rewrite this paragraph, copied from first2years paper} One might naively expect that self\nobreakdashes-consistency would require the 90\% confidence area and searched area histograms to intersect at 90\% of detections, but this is not generally required because the posteriors of different events have widely different dimensions. However, it is true that 90\% of sources should be found within their respective 90\% confidence contours. This can be formalized into a graphical self\nobreakdashes-consistency test called a probability\nobreakdashes--probability ($P$\nobreakdashes--$P$) plot (see \citealt{SiderySkyLocalizationComparison} for applications in \ac{GW} parameter estimation). For each event, one follows the water\nobreakdashes-filling procedure to walk outward from the most probable pixel until one reaches the pixel that contains the true sky location, noting the total probability $p$ that has been traversed.

A cumulative histogram of $p$ should be diagonal, within a binomial confidence band. It is already well established that LALINFERENCE localizations satisfy the $P$\nobreakdashes--$P$ plot test when deployed with accurate templates and reasonable priors. We found at first that \ac{BAYESTAR}'s $P$\nobreakdashes--$P$ plots tended to sag below the diagonal, indicating that though the accuracy (i.e., searched area) was comparable to the stochastic samplers, the \emph{precision} was overstated with confidence intervals that were only about 70\% of the correct area. This was rectified by pre\nobreakdashes-scaling the \acp{SNR} from GSTLAL\_INSPIRAL by a factor of 0.83 prior to running \ac{BAYESTAR}. This correction factor suggests that, for example, an \ac{SNR} 10 trigger from GSTLAL\_INSPIRAL has the \emph{effective information content} of an \ac{SNR} 8.3 signal. The missing information may be due to losses from the discreteness of the template bank, from the \ac{SVD}, from mismatch between the matched\nobreakdashes-filter templates and the simulated signals, from the small but nonzero correlations between masses and intrinsic parameters, or from elsewhere within the detection pipeline. The correction is hard\nobreakdashes-coded into the rapid localization. With it, the $P$\nobreakdashes--$P$ plots are diagonalized without negatively affecting the searched area (see Figure~\ref{fig:pp}).

@@

Several features are apparent. First, at any number of threads, the two configurations have similar run times, although the 2016 events contain a subpopulation of outliers that take about 2.5 times as long as the 2015 events. These are probably due to taking one of the more expensive code branches in the distance integral interpolation. Second, the run times decrease proportionally to the number of threads. A slight flattening at 32 threads may be due to increasing use of hyperthreading. Based on experiences running \ac{BAYESTAR} on the 32\nobreakdashes-core (64 threads with hyperthreading) cluster login machine, we expect the almost ideal parallel speedup to continue on machines with more threads. With just one thread, the \ac{BAYESTAR} analysis takes 100\nobreakdashes--400~s, already orders of magnitude faster than the full parameter estimation. With 32 threads, \ac{BAYESTAR} takes just 20\nobreakdashes--60~s. This latency is comparable to the other stages (data aggregation, trigger generation, alert distribution) in the real\nobreakdashes-time \ac{BNS} analysis. The 32\nobreakdashes-thread configuration is representative of how \ac{BAYESTAR} might be deployed in early Advanced \ac{LIGO}.\footnote{\ac{BAYESTAR} has been successfully ported to the Intel's Many Integrated Core~(MIC) architecture and has been tested in a 500 thread configuration on a system with dual Intel Xeon Phi 5110P coprocessors.}

>>>>

\subsection{Integral over distance}
\label{sec:distance}

The distance integral is now performed differently from what we initially described in \cite{leo-singer-thesis,BerryLocalization}; the method described in the present work is about an order of magnitude faster. The innermost integral over distance $r$ may be written as
%
\begin{align}\label{eq:distance-integral}
    \mathscr{F} &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}\right]
        I_0\left[\frac{b}{r}\right] r^m dr \nonumber \\
    &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}+\frac{b}{r}\right]
        \overline{I}_0\left[\frac{b}{r}\right] r^m dr
\end{align}
%
or, completing the square,
%
\begin{align}
    \mathscr{F} &= \exp\left[\frac{p^2}{{r_0}^2}\right]
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\left(\frac{p}{r} - \frac{p}{r_0}\right)^2\right]
        \overline{I}_0\left[\frac{2p^2}{r r_0}\right] r^m dr \\
    &= \exp\left[\frac{p^2}{{r_0}^2}\right] \mathscr{G},
\end{align}
%
where
%
\begin{align}
    p^2 &= \frac{1}{2} r^2 \sum_i {\rho_i}^2 \label{eq:likelihood-p-factor} \\
    b &= r \left| \sum_i \hat\rho_i \rho_i e^{i \tilde{\gamma}_i}
        a_i^*(\tilde{\tau}_i) \right| \label{eq:likelihood-b-factor} \\
    r_0 &= 2 p^2 / b \\
    \overline{I}_0(x) &= \exp(-|x|) I_0(x).
\end{align}
%
The coefficients $p^2$ and $b$ are nonnegative and independent of distance. $p$ has a maximum value of
%
\begin{equation}
    p_\mathrm{max} = \sqrt{\frac{1}{2} \sum_i \left(\frac{r_{1,i}}{r}\right)^2}.
\end{equation}
%
The symbol $\overline{I}_0$ denotes an exponentially scaled Bessel function. In the limit of large argument, $I_0(|x|) \sim \exp(|x|) / \sqrt{2 \pi |x|}$ \citep{Olver:2010:NHMF,NIST:DLMF}\footnote{\url{http://dlmf.nist.gov/10.40.E1}}. The scaled Bessel is useful for evaluation on a computer because it has a relatively small range $(0, 1]$ and varies slowly in proportion to $x^{1/2}$.

@@

\section{Parallelization}

\ac{MCMC} and similar stochastic schemes are typically very resistant to parallelization. However, \ac{BAYESTAR} is completely deterministic and easily parallelizable because each pixel can be evaluated independently from all of the others. \ac{BAYESTAR} consists of nine computationally intensive loops: the generation of the distance integral lookup table and the eight loops over pixels in the adaptive HEALPix sampling step. The iterations of each loop are distributed across multiple cores using OpenMP\footnote{http://openmp.org/} compiler directives. In Section~\ref{sec:run-time}, we will show that \ac{BAYESTAR}'s run time is almost perfectly proportional to the number of cores, demonstrating that the serial sections (the sorts between the adaptation steps) are a negligible contribution to the overall wall clock time.

@@

We measured sky localization areas for each event as follows. First, we ranked the \ac{HEALPix} pixels by descending posterior probability. Then, we computed the cumulative sum of the pixels in that order. Finally, we searched for the index of the pixel whose cumulative sum was equal to a given value: for example, 0.9 if we are interested in the 90\% credible area. That pixel index times the area per pixel is the area of the smallest region of the specified credible level. This area can be thought of measuring the precision of the sky localization: it is a measure of the scale of the posterior distribution.

We can construct a second measure, called the searched area, as the smallest such constructed area that contains the true location of the source. A telescope with a \ac{FOV} that is small compared to the characteristic scale of the posterior would intercept the true location of the source after covering the searched area. This measure is mainly useful because it measures the accuracy of the sky localization independently of the precision. In other words, it treats the sky map as merely a ranking statistic.

@@

\subsection{Self consistency}

As we observed above, the area of a given credible region describes the precision of the sky localization, whereas the searched area describes the accuracy. However, self consistency requires that the two are related. For example, we should find that on average 90\% of events have their true locations contained within their respective 90\% credible regions. More generally, if we make a cumulative histogram of the credible levels corresponding to the searched areas of all of the events, then we should obtain a diagonal line (with small deviations due to finite sample size). This test, popularized for \ac{GW} data analysis by \cite{SiderySkyLocalizationComparison}, is a necessary but not sufficient condition for the validity of any Bayesian parameter estimation scheme.

It is already well established that LALINFERENCE localizations satisfy the $P$\nobreakdashes--$P$ plot test when deployed with accurate templates and reasonable priors. We found at first that \ac{BAYESTAR}'s $P$\nobreakdashes--$P$ plots tended to sag below the diagonal, indicating that though the accuracy (i.e., searched area) was comparable to LALINFERENCE, the \emph{precision} was overstated, with confidence intervals that were only about 70\% of the correct area. This was rectified by pre\nobreakdashes-scaling the \acp{SNR} from GSTLAL\_INSPIRAL by a factor of 0.83 prior to running \ac{BAYESTAR}. This correction factor suggests that, for example, an \ac{SNR} 10 trigger from GSTLAL\_INSPIRAL has the \emph{effective information content} of an \ac{SNR} 8.3 signal. The missing information may be due to losses from the discreteness of the template bank, from the \ac{SVD}, from mismatch between the matched\nobreakdashes-filter templates and the simulated signals, from the small but nonzero correlations between masses and intrinsic parameters, or from elsewhere within the detection pipeline. The correction is hard\nobreakdashes-coded into the rapid localization. With it, the $P$\nobreakdashes--$P$ plots are diagonalized without negatively affecting the searched area (see Figure~\ref{fig:pp}).

@@

Several features are apparent. First, at any number of threads, the two configurations have similar run times, although the 2016 events contain a subpopulation of outliers that take about 2.5 times as long as the 2015 events. These are probably due to taking one of the more expensive code branches in the distance integral interpolation. Second, the run times decrease proportionally to the number of threads. A slight flattening at 32 threads may be due to increasing use of hyperthreading. Based on experiences running \ac{BAYESTAR} on the 32\nobreakdashes-core (64 threads with hyperthreading) cluster login machine, we expect the almost ideal parallel speedup to continue on machines with more threads.

With just one thread, the \ac{BAYESTAR} analysis takes 87\nobreakdashes--384~s, already orders of magnitude faster than the full parameter estimation. With 32 threads, \ac{BAYESTAR} takes just 4\nobreakdashes--15~s. This latency is comparable to the other stages (data aggregation, trigger generation, alert distribution) in the real\nobreakdashes-time \ac{BNS} analysis; therefore any significant further speedup would require significant changes to all of the analyses preceeding \ac{BAYESTAR}. The 32\nobreakdashes-thread configuration is representative of how \ac{BAYESTAR} might be deployed in early Advanced \ac{LIGO}.\footnote{\ac{BAYESTAR} has been successfully ported to the Intel's Many Integrated Core~(MIC) architecture and has been tested in a 500 thread configuration on a system with dual Intel Xeon Phi 5110P coprocessors.} For comparison, sky localization with LALINFERENCE takes about 100 hours \cite{BerryLocalization}.

Note that this benchmark shows \ac{BAYESTAR} to be an order of magnitude faster than what was reported in \cite{leo-singer-thesis,BerryLocalization} due to the changes in the distance integration scheme that we noted in Section~\ref{sec:distance}.
